Improvement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-06-08 rigid_bodies.tex
Change command from modified to refined

\usepackage{xr}
\externaldocument{rigid_bodies_supp_info}
\usepackage{booktabs}
\usepackage{threeparttable}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[T1]{fontenc}
\usepackage[table]{xcolor}
\definecolor{lightgray}{gray}{0.9}
\usepackage{tikz}
\usetikzlibrary{decorations.text}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}   % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                % transposition
\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville
\newcommand{\timestep}{h}
\newcommand{\modified}[1]{\widetilde{#1}}

@@

As demonstrated in Appendix~\ref{sec:rigid body shadow hamiltonian}, the \textit{refined} Hamiltonian for a Verlet-type method applied to a system of rigid bodies is given by $\modified{\Ham{}} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are \textit{refined} kinetic and potential energies.
The latter is given by
\begin{equation}
\label{eq:modified potential energy}
\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
\end{equation}
Note that $\modified U$, such as $U$, is independent of $\vt p$ and $\vt \pi$.
In turn, the \textit{refined} kinetic energy can be expressed as
\begin{equation}
\label{eq:modified kinetic energy}
\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\modified{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, given by $\tilde{\mt \Omega} = {\mt \Omega} + {\timestep^2}/{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)$, where
\begin{equation*}
{\mt \Omega} = \left[\begin{array}{cc}
{\mt M}^{-1} & \mt 0 \\
\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
\end{array}\right]
\end{equation*}
and ${\mt U}_{\vt x \vt x}$ is the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, given by
\begin{equation*}
{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
U_{\vt r \vt r} & U_{\vt r \vt q} \\
U_{\vt q \vt r} & U_{\vt q \vt q}
\end{array}\right].
\end{equation*}

Hence, the trajectory generated by the splitting scheme of Eq.~\ref{eq:splitting NVE}, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will approach even more closely the solution of a modified ODE system given by
\begin{equation*}
\dot{\vt r} = \diff{\modified K}{\vt p}, \;
\dot{\vt q} = \diff{\modified K}{\vt \pi}, \;
\dot{\vt p} = -\diff{\modified{\Ham{}}}{\vt r}, \; \text{and} \;
\dot{\vt \pi} = -\diff{\modified{\Ham{}}}{\vt q}.
\end{equation*}

The first two of these equations can be combined together so that we can use Eq.~\eqref{eq:modified kinetic energy} to obtain
\begin{equation}
\label{eq:shadow ODE system}
\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] = \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right].
\end{equation}

Analytical calculation of $\modified U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
In contrast, although it is possible to compute $\modified K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
Fortunately, we can estimate it rather easily though a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
\begin{equation}
\label{eq:modified kinetic energy estimator}
\modified K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(h^n).
\end{equation}

@@

\subsection{Adjustment of the NVT Dynamics}
\label{sec:refined_method}
Bringing the attention back to the NVT case, our proposal consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, produces a phase-space move that is more consistent with the modified Hamiltonian ODE system than with the original one.
Thus, we should adjust the Nos\'e-Hoover chain so as to provide a distribution proportional to $e^{-\beta \modified{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting scheme, as described before.
The only required modification is that $\modified K$ replaces $K$ in Eq.~\eqref{eq:generalized force NVT}, which defines the force $G_1$.
By contrasting such definition with Eqs.~\eqref{eq:modified kinetic energy} and \eqref{eq:shadow ODE system}, we observe that $G_1$ becomes
\begin{equation}
\label{eq:thermostat_force}
G_1 = 2 \modified K - L k_\text{\tiny B} T.
\end{equation}
Also, the ``extended energy'' preserved by this approach is now given by
\begin{equation}
\label{eq:this extended energy}
\widetilde{H} = \modified K + \modified U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
\end{equation}

One instant in which $\modified K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
Notice that such action takes place only once per time step, intercalating with non-Hamiltonian propagations.
Therefore, there is no record beforehand of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$ with high accuracy.
This could be solved by executing a number of virtual Verlet steps, both forward and backward in time, so as to generate a virtual Hamiltonian trajectory that coincides with the actual trajectory in the two phase-space points which are input and output of the central propagator in Eq.~\eqref{eq:splitting NVT}.
Nonetheless, this might require additional force computations per actual time step, thus increasing the cost of the simulation.
Fortunately, only positions and orientations are required for the numerical differentiation.
This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time, as schematically illustrated in Fig.~\ref{fig:virtual}.
By incomplete we mean that the step is aborted once the positions and orientations have been updated.
\begin{figure}
\begin{tikzpicture}
\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{$r(t)$};
\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
\node at (0.0,-0.2)[below]{$r_{t-h}$};
\node at (2.2,-0.2)[below]{$r_{t}$};
\node at (4.4,-0.2)[below]{$r_{t+h}$};
\node at (6.6,-0.2)[below]{$r_{t+2h}$};
\path[->,thick] (1) edge [bend left=45] (2);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
\path[dashed,->,thick] (1) edge [bend right=45] (0);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
\path[dashed,->,thick] (2) edge [bend left=45] (3);
\end{tikzpicture}
\caption{Schematic representation}
\label{fig:virtual}
\end{figure}

In addition, $\modified K$ must be reevaluated after the action of propagator $e^{\frac{\timestep}{2n} {\mathcal L}_0}$, which scales the momentum vectors $\vt p$ and $\vt \pi$ by a common factor $e^{-\alpha_1^\ast \frac{\timestep}{2n}}$.
Because such propagator leaves the positions and quaternions intact and, as a consequence, the matrix $\modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\modified K$ goes as simply as ${\modified K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\modified K}^\ast$.

Please note that Eq.~\ref{eq:generalized force NVT} is closely related to Tolman's generalized equipartition theorem \cite{Uline_2008}, which states that
\begin{equation*}
\left\langle \tr{\vt p} \diff{\modified{\Ham{}}}{\vt p} + \tr{\vt \pi} \diff{\modified{\Ham{}}}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T.
\end{equation*}
By substituting $\modified{\Ham{}}$ in this equation, we obtain that
\begin{equation}
2\left\langle \modified{K} \right\rangle = (6N - 3) k_\text{\tiny B} T.
\end{equation}
This result, with the \textit{refined} kinetic energy between brackets, provides a generalization for $h > 0$ of the simulated temperature \cite{Eastwood_2010}.

Finally, given that configurations $x_i$ are sampled from a distribution proportional to $\exp(-\beta \modified{\Ham{}})$, we need to reweight the computed observables $A(x_i)$ to recover the original distribution $\exp (-\beta \Ham{})$.
This is done as follows:
\begin{equation}
\label{eq:reweighting}
\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\modified{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\modified{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
\end{equation}

@@

In Fig.~\ref{fig:nve} we show the time evolution of both $\modified{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$). For this, we used the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit solution of rotation. As expected, $\modified{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ = 1 fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$. 
It should be remarked that $\modified{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap. As a result of such approximation, we have larger fluctuations in $\modified{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).

\begin{figure}
	\includegraphics{Figures/NVE.eps}
	\caption{Total energy $\Ham{}$ (solid lines) and \textit{refined} Hamiltonian $\modified{\Ham{}}$ (dotted lines) for different time-step sizes ($\timestep$) obtained from NVE MD simulations of 903 TIP3P\cite{Jorgensen_1983} water molecules using the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit rotation method.}
	\label{fig:nve}
\end{figure}

@@

Here we define the energy drift, denoted by $R$, as the the rate of change of energy with time, which corresponds to the angular coefficient obtained from linear regression fitting.
Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\ref{eq:this extended energy}) for our method, which from now on will be referred as ``This work''.
On the other hand, the symmetric scheme due to Martyna \textit{et al.} \cite{Martyna_1996} should preserve $H$ (Eq.~\ref{eq:nvt extended energy}).
In Fig.~\ref{fig:energy_drift}, we present the computed $R$ as a function of the time-step size, and one can see that for these numerical schemes the energy starts to drift upwards from $h$ = 5 fs. 
Also, the long-term stability is only modestly improved by our method. Evidently, the impact of the lack of symmetry on $R$, if any, is negligible.
Fig.~\ref{fig:energy_drift} also contains the drift in $\modified{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of our method. It is worth mentioning that the stability threshold for the NVE method corresponds to $h$ = 6 fs.
The NVT methods discussed so far allow to extend that limit to 10 fs (see Table~\ref{table:stability} in the Supplementary Material). As one can notice in Fig.~\ref{fig:energy_drift}, the drift in the \textit{refined} Hamiltonian part for our NVT method (squares) remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration.
This has already been noticed by Davidchack\cite{Davidchack_2009}, and previously by Bond \textit{et al.} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical error''.

\begin{figure}
	\includegraphics{Figures/energy_drift.eps}
    \caption{Influence of the time-step size ($h$) on the long-time drift ($R$) in: the extended energy ($\modified{H}$, down triangles) and its \textit{refined} Hamiltonian ($\modified{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the \textit{refined} Hamiltonian ($\modified{\Ham{}}$, circles) for the numerical scheme of Eq.~\ref{eq:splitting NVE} with unsplit solution of rotation.}
	\label{fig:energy_drift}
\end{figure}

@@

With the aim of highlighting the importance of reweighting to recover averages of the physical system, in Table~\ref{table:reweight} we present the calculated temperature and potential energy for each $\timestep$ before and after applying Eq.~\ref{eq:reweighting} for time-step sizes smaller 7 fs, as we saw in Fig. that for ths case, the temperature does not reproduce the specified temperature.
The ``refined'' temperature $\modified{T}$ was calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
Of course, as $\modified{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the specified temperature.
Indeed, for all the time-step sizes considered the agreement is excellent.
We see that in the absence of reweighting, the computed temperatures are notably influenced by $h$.
Also note that a small, still clearly influencing systematic error is present in our scheme, as both temperature and potential energy are slightly modified by $h$. This could, in principle, be remediated by using a higher-order ``refined'' Hamiltonian.

In addition, we are able to ``measure'' how far our sample is from the state of interest with physical Hamiltonian $\modified{\Ham{}}$ by calculating the free energy differences ($\Delta F$) between the sampled state, whose Hamiltonian is given by $\modified{\Ham{}}$, and that state of interest.
$\Delta F$ corresponds to the denominator of Eq.~\ref{eq:reweighting}.
As expected, from Table~\ref{table:reweight} one verifies that $\Delta F$ is only negligible for $h$ = 1 fs and systematically increases with $h$.

>>>>

\usepackage{xr}
\externaldocument{rigid_bodies_supp_info}
\usepackage{booktabs}
\usepackage{threeparttable}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[T1]{fontenc}
\usepackage[table]{xcolor}
\definecolor{lightgray}{gray}{0.9}
\usepackage{tikz}
\usetikzlibrary{decorations.text}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}   % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                % transposition
\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville
\newcommand{\timestep}{h}
\newcommand{\refined}[1]{\widetilde{#1}}

@@

As demonstrated in Appendix~\ref{sec:rigid body shadow hamiltonian}, the \textit{refined} Hamiltonian for a Verlet-type method applied to a system of rigid bodies is given by $\refined{\Ham{}} = \refined K + \refined U + \mathcal{O}(h^4)$, where $\refined K$ and $\refined U$ are \textit{refined} kinetic and potential energies.
The latter is given by
\begin{equation}
\label{eq:modified potential energy}
\refined U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
\end{equation}
Note that $\refined U$, such as $U$, is independent of $\vt p$ and $\vt \pi$.
In turn, the \textit{refined} kinetic energy can be expressed as
\begin{equation}
\label{eq:modified kinetic energy}
\refined K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\refined{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, given by $\tilde{\mt \Omega} = {\mt \Omega} + {\timestep^2}/{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)$, where
\begin{equation*}
{\mt \Omega} = \left[\begin{array}{cc}
{\mt M}^{-1} & \mt 0 \\
\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
\end{array}\right]
\end{equation*}
and ${\mt U}_{\vt x \vt x}$ is the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, given by
\begin{equation*}
{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
U_{\vt r \vt r} & U_{\vt r \vt q} \\
U_{\vt q \vt r} & U_{\vt q \vt q}
\end{array}\right].
\end{equation*}

Hence, the trajectory generated by the splitting scheme of Eq.~\ref{eq:splitting NVE}, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will approach even more closely the solution of a modified ODE system given by
\begin{equation*}
\dot{\vt r} = \diff{\refined K}{\vt p}, \;
\dot{\vt q} = \diff{\refined K}{\vt \pi}, \;
\dot{\vt p} = -\diff{\refined{\Ham{}}}{\vt r}, \; \text{and} \;
\dot{\vt \pi} = -\diff{\refined{\Ham{}}}{\vt q}.
\end{equation*}

The first two of these equations can be combined together so that we can use Eq.~\eqref{eq:modified kinetic energy} to obtain
\begin{equation}
\label{eq:shadow ODE system}
\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] = \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right].
\end{equation}

Analytical calculation of $\refined U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
In contrast, although it is possible to compute $\refined K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
Fortunately, we can estimate it rather easily though a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
\begin{equation}
\label{eq:modified kinetic energy estimator}
\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(h^n).
\end{equation}

@@

\subsection{Adjustment of the NVT Dynamics}
\label{sec:refined_method}
Bringing the attention back to the NVT case, our proposal consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, produces a phase-space move that is more consistent with the modified Hamiltonian ODE system than with the original one.
Thus, we should adjust the Nos\'e-Hoover chain so as to provide a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting scheme, as described before.
The only required modification is that $\refined K$ replaces $K$ in Eq.~\eqref{eq:generalized force NVT}, which defines the force $G_1$.
By contrasting such definition with Eqs.~\eqref{eq:modified kinetic energy} and \eqref{eq:shadow ODE system}, we observe that $G_1$ becomes
\begin{equation}
\label{eq:thermostat_force}
G_1 = 2 \refined K - L k_\text{\tiny B} T.
\end{equation}
Also, the ``extended energy'' preserved by this approach is now given by
\begin{equation}
\label{eq:this extended energy}
\widetilde{H} = \refined K + \refined U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
\end{equation}

One instant in which $\refined K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
Notice that such action takes place only once per time step, intercalating with non-Hamiltonian propagations.
Therefore, there is no record beforehand of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$ with high accuracy.
This could be solved by executing a number of virtual Verlet steps, both forward and backward in time, so as to generate a virtual Hamiltonian trajectory that coincides with the actual trajectory in the two phase-space points which are input and output of the central propagator in Eq.~\eqref{eq:splitting NVT}.
Nonetheless, this might require additional force computations per actual time step, thus increasing the cost of the simulation.
Fortunately, only positions and orientations are required for the numerical differentiation.
This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time, as schematically illustrated in Fig.~\ref{fig:virtual}.
By incomplete we mean that the step is aborted once the positions and orientations have been updated.
\begin{figure}
\begin{tikzpicture}
\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{$r(t)$};
\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
\node at (0.0,-0.2)[below]{$r_{t-h}$};
\node at (2.2,-0.2)[below]{$r_{t}$};
\node at (4.4,-0.2)[below]{$r_{t+h}$};
\node at (6.6,-0.2)[below]{$r_{t+2h}$};
\path[->,thick] (1) edge [bend left=45] (2);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
\path[dashed,->,thick] (1) edge [bend right=45] (0);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
\path[dashed,->,thick] (2) edge [bend left=45] (3);
\end{tikzpicture}
\caption{Schematic representation}
\label{fig:virtual}
\end{figure}

In addition, $\refined K$ must be reevaluated after the action of propagator $e^{\frac{\timestep}{2n} {\mathcal L}_0}$, which scales the momentum vectors $\vt p$ and $\vt \pi$ by a common factor $e^{-\alpha_1^\ast \frac{\timestep}{2n}}$.
Because such propagator leaves the positions and quaternions intact and, as a consequence, the matrix $\refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\refined K$ goes as simply as ${\refined K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\refined K}^\ast$.

Please note that Eq.~\ref{eq:generalized force NVT} is closely related to Tolman's generalized equipartition theorem \cite{Uline_2008}, which states that
\begin{equation*}
\left\langle \tr{\vt p} \diff{\refined{\Ham{}}}{\vt p} + \tr{\vt \pi} \diff{\refined{\Ham{}}}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T.
\end{equation*}
By substituting $\refined{\Ham{}}$ in this equation, we obtain that
\begin{equation}
2\left\langle \refined{K} \right\rangle = (6N - 3) k_\text{\tiny B} T.
\end{equation}
This result, with the \textit{refined} kinetic energy between brackets, provides a generalization for $h > 0$ of the simulated temperature \cite{Eastwood_2010}.

Finally, given that configurations $x_i$ are sampled from a distribution proportional to $\exp(-\beta \refined{\Ham{}})$, we need to reweight the computed observables $A(x_i)$ to recover the original distribution $\exp (-\beta \Ham{})$.
This is done as follows:
\begin{equation}
\label{eq:reweighting}
\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
\end{equation}

@@

In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$). For this, we used the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit solution of rotation. As expected, $\refined{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ = 1 fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$. 
It should be remarked that $\refined{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap. As a result of such approximation, we have larger fluctuations in $\refined{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).

\begin{figure}
	\includegraphics{Figures/NVE.eps}
	\caption{Total energy $\Ham{}$ (solid lines) and \textit{refined} Hamiltonian $\refined{\Ham{}}$ (dotted lines) for different time-step sizes ($\timestep$) obtained from NVE MD simulations of 903 TIP3P\cite{Jorgensen_1983} water molecules using the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit rotation method.}
	\label{fig:nve}
\end{figure}

@@

Here we define the energy drift, denoted by $R$, as the the rate of change of energy with time, which corresponds to the angular coefficient obtained from linear regression fitting.
Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\ref{eq:this extended energy}) for our method, which from now on will be referred as ``This work''.
On the other hand, the symmetric scheme due to Martyna \textit{et al.} \cite{Martyna_1996} should preserve $H$ (Eq.~\ref{eq:nvt extended energy}).
In Fig.~\ref{fig:energy_drift}, we present the computed $R$ as a function of the time-step size, and one can see that for these numerical schemes the energy starts to drift upwards from $h$ = 5 fs. 
Also, the long-term stability is only modestly improved by our method. Evidently, the impact of the lack of symmetry on $R$, if any, is negligible.
Fig.~\ref{fig:energy_drift} also contains the drift in $\refined{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of our method. It is worth mentioning that the stability threshold for the NVE method corresponds to $h$ = 6 fs.
The NVT methods discussed so far allow to extend that limit to 10 fs (see Table~\ref{table:stability} in the Supplementary Material). As one can notice in Fig.~\ref{fig:energy_drift}, the drift in the \textit{refined} Hamiltonian part for our NVT method (squares) remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration.
This has already been noticed by Davidchack\cite{Davidchack_2009}, and previously by Bond \textit{et al.} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical error''.

\begin{figure}
	\includegraphics{Figures/energy_drift.eps}
    \caption{Influence of the time-step size ($h$) on the long-time drift ($R$) in: the extended energy ($\refined{H}$, down triangles) and its \textit{refined} Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the \textit{refined} Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\ref{eq:splitting NVE} with unsplit solution of rotation.}
	\label{fig:energy_drift}
\end{figure}

@@

With the aim of highlighting the importance of reweighting to recover averages of the physical system, in Table~\ref{table:reweight} we present the calculated temperature and potential energy for each $\timestep$ before and after applying Eq.~\ref{eq:reweighting} for time-step sizes smaller 7 fs, as we saw in Fig. that for ths case, the temperature does not reproduce the specified temperature.
The ``refined'' temperature $\refined{T}$ was calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the specified temperature.
Indeed, for all the time-step sizes considered the agreement is excellent.
We see that in the absence of reweighting, the computed temperatures are notably influenced by $h$.
Also note that a small, still clearly influencing systematic error is present in our scheme, as both temperature and potential energy are slightly modified by $h$. This could, in principle, be remediated by using a higher-order ``refined'' Hamiltonian.

In addition, we are able to ``measure'' how far our sample is from the state of interest with physical Hamiltonian $\refined{\Ham{}}$ by calculating the free energy differences ($\Delta F$) between the sampled state, whose Hamiltonian is given by $\refined{\Ham{}}$, and that state of interest.
$\Delta F$ corresponds to the denominator of Eq.~\ref{eq:reweighting}.
As expected, from Table~\ref{table:reweight} one verifies that $\Delta F$ is only negligible for $h$ = 1 fs and systematically increases with $h$.

\begin{table*}
	\caption{Effect of the time-step size on .... of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes. uncertainties in free energy never exced 1e-4}
    \label{table:reweight}
    \sisetup{
table-number-alignment = center
}
    \begin{tabular}{SS[separate-uncertainty,table-figures-uncertainty = 2]S[separate-uncertainty,table-figures-uncertainty=2]S[separate-uncertainty,table-figures-uncertainty= 1,table-figures-decimal=4]S[separate-uncertainty,table-figures-uncertainty= 1]S[separate-uncertainty,table-figures-uncertainty= 1,table-figures-decimal=4]S}
     & &  \multicolumn{2}{c}{no reweighting} & \multicolumn{2}{c}{after reweighting} & \\
     \cmidrule[0.5mm](lr){3-4} \cmidrule[0.5mm](lr){5-6}
     $\timestep$ & $\refined{T}$  & $T$  & {$\langle U/N \rangle$}  & $T$ & {$\langle U/N \rangle$}   & {$\Delta F/N$} \\
        $\text{(fs)}$  & $\text{(K)}$ & $\text{(K)}$ & $\text{(kcal/mol)}$  & $\text{(K)}$ & $\text{(kcal/mol)}$ & $\text{(kcal/mol)}$ \\
	1 & 298.00(1) & 297.71(1) & -9.1015(3) & 298.00(1) & -9.1028(3) & -0.0006 \\
	2 & 297.99(1) & 296.82(1) & -9.0970(3)  & 297.98(1)  & -9.1024(3) & -0.004 \\
	3 & 297.98(1) & 295.38(1) & -9.0885(3) & 297.92(1) & -9.1005(3) & -0.008 \\
	4 & 298.01(1) & 293.49(1)& -9.0764(3) & 297.85(3) & -9.0969(3) & -0.013 \\
	5 & 297.99(1)& 291.13(1)& -9.0577(3) & 297.6(2) & -9.091(1) & -0.033 \\
    6 & 298.00(1) & 288.46(1) & -9.0314(3) & 296.8(3) & -9.075(3) &  -0.026 \\
	7 & 297.99(1) & 285.57(1) & -8.9944(3)  & 295.0(5)  & -9.040(4) & -0.056 \\
	\end{tabular}
\end{table*}
